---
title: "Homework 03"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

1. Write a function to drop URLs from a message body. The function should take a message body as its input and return the content with URLs removed. Apply the function to the sample text testmsgBody in HW3Setup.R to show that the function works.  

*********************************************************Homework code********************************************************************************
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# The following setup code is based on the code for Chapter 3 of 
# 
# Nolan, Deborah and Temple Lang, Duncan. Data Science in R: A Case Studies Approach to 
# Computational Reasoning and Problem Solving. CRC Press, 2015. 
# http://rdatasciencecases.org/

# The following lines from sections 3.3 through 3.6.3 define functions and objects 
# that are needed or may be useful for the exercises.
spamPath = "~/Stat480/RDataScience/SpamAssassinMessages"

dirNames = list.files(path = paste(spamPath, "messages", 
                                   sep = .Platform$file.sep))
fullDirNames = paste(spamPath, "messages", dirNames, 
                     sep = .Platform$file.sep)

indx = c(1:5, 15, 27, 68, 69, 329, 404, 427, 516, 852, 971)
fn = list.files(fullDirNames[1], full.names = TRUE)[indx]
sampleEmail = sapply(fn, readLines)

splitMessage = function(msg) {
  splitPoint = match("", msg)
  header = msg[1:(splitPoint-1)]
  body = msg[ -(1:splitPoint) ]
  return(list(header = header, body = body))
}

sampleSplit = lapply(sampleEmail, splitMessage)

getBoundary = function(header) {
  boundaryIdx = grep("boundary=", header)
  boundary = gsub('"', "", header[boundaryIdx])
  gsub(".*boundary= *([^;]*);?.*", "\\1", boundary)
}

dropAttach = function(body, boundary){
  
  bString = paste("--", boundary, sep = "")
  bStringLocs = which(bString == body)
  
  # if there are fewer than 2 beginning boundary strings, 
  # there is on attachment to drop
  if (length(bStringLocs) <= 1) return(body)
  
  # do ending string processing
  eString = paste("--", boundary, "--", sep = "")
  eStringLoc = which(eString == body)
  
  # if no ending boundary string, grab contents between the first 
  # two beginning boundary strings as the message body
  if (length(eStringLoc) == 0) 
    return(body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1)])
  
  # typical case of well-formed email with attachments
  # grab contents between first two beginning boundary strings and 
  # add lines after ending boundary string
  n = length(body)
  if (eStringLoc < n) 
    return( body[ c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), 
                     ( (eStringLoc + 1) : n )) ] )
  
  # fall through case
  # note that the result is the same as the 
  # length(eStringLoc) == 0 case, so code could be simplified by 
  # dropping that case and modifying the eStringLoc < n check to 
  # be 0 < eStringLoc < n
  return( body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1) ])
}

library(tm)
stopWords = stopwords()
cleanSW = tolower(gsub("[[:punct:]0-9[:blank:]]+", " ", stopWords))
SWords = unlist(strsplit(cleanSW, "[[:blank:]]+"))
SWords = SWords[ nchar(SWords) > 1 ]
stopWords = unique(SWords)

cleanText =
  function(msg)   {
    tolower(gsub("[[:punct:]0-9[:space:][:blank:]]+", " ", msg))
  }

findMsgWords = 
  function(msg, stopWords) {
    if(is.null(msg))
      return(character())
    
    words = unique(unlist(strsplit(cleanText(msg), "[[:blank:]\t]+")))
    
    # drop empty and 1 letter words
    words = words[ nchar(words) > 1]
    words = words[ !( words %in% stopWords) ]
    invisible(words)
  }

processAllWords = function(dirName, stopWords)
{
  # read all files in the directory
  fileNames = list.files(dirName, full.names = TRUE)
  # drop files that are not email, i.e., cmds
  notEmail = grep("cmds$", fileNames)
  if ( length(notEmail) > 0) fileNames = fileNames[ - notEmail ]
  
  messages = lapply(fileNames, readLines, encoding = "latin1")
  
  # split header and body
  emailSplit = lapply(messages, splitMessage)
  # put body and header in own lists
  bodyList = lapply(emailSplit, function(msg) msg$body)
  headerList = lapply(emailSplit, function(msg) msg$header)
  rm(emailSplit)
  
  # determine which messages have attachments
  hasAttach = sapply(headerList, function(header) {
    CTloc = grep("Content-Type", header)
    if (length(CTloc) == 0) return(0)
    multi = grep("multi", tolower(header[CTloc])) 
    if (length(multi) == 0) return(0)
    multi
  })
  
  hasAttach = which(hasAttach > 0)
  
  # find boundary strings for messages with attachments
  boundaries = sapply(headerList[hasAttach], getBoundary)
  
  # drop attachments from message body
  bodyList[hasAttach] = mapply(dropAttach, bodyList[hasAttach], 
                               boundaries, SIMPLIFY = FALSE)
  
  # extract words from body
  msgWordsList = lapply(bodyList, findMsgWords, stopWords)
  
  invisible(msgWordsList)
}

msgWordsList = lapply(fullDirNames, processAllWords, 
                      stopWords = stopWords) 

numMsgs = sapply(msgWordsList, length)

isSpam = rep(c(FALSE, FALSE, FALSE, TRUE, TRUE), numMsgs)

msgWordsList = unlist(msgWordsList, recursive = FALSE)

# Determine number of spam and ham messages for sampling.
numEmail = length(isSpam)
numSpam = sum(isSpam)
numHam = numEmail - numSpam

# Set a particular seed, so the results will be reproducible.
set.seed(418910)

# Take approximately 1/3 of the spam and ham messages as our test spam and ham messages.
testSpamIdx = sample(numSpam, size = floor(numSpam/3))
testHamIdx = sample(numHam, size = floor(numHam/3))

# Use the test indices to select word lists for test messages.
# Use training indices to select word lists for training messages.
testMsgWords = c((msgWordsList[isSpam])[testSpamIdx],
                 (msgWordsList[!isSpam])[testHamIdx] )
trainMsgWords = c((msgWordsList[isSpam])[ - testSpamIdx], 
                  (msgWordsList[!isSpam])[ - testHamIdx])

# Create variables indicating which testing and training messages are spam and not.
testIsSpam = rep(c(TRUE, FALSE), 
                 c(length(testSpamIdx), length(testHamIdx)))
trainIsSpam = rep(c(TRUE, FALSE), 
                  c(numSpam - length(testSpamIdx), 
                    numHam - length(testHamIdx)))


computeFreqs =
  function(wordsList, spam, bow = unique(unlist(wordsList)))
  {
    # create a matrix for spam, ham, and log odds
    wordTable = matrix(0.5, nrow = 4, ncol = length(bow), 
                       dimnames = list(c("spam", "ham", 
                                         "presentLogOdds", 
                                         "absentLogOdds"),  bow))
    
    # For each spam message, add 1/2 to counts for words in message
    counts.spam = table(unlist(lapply(wordsList[spam], unique)))
    wordTable["spam", names(counts.spam)] = counts.spam + .5
    
    # Similarly for ham messages
    counts.ham = table(unlist(lapply(wordsList[!spam], unique)))  
    wordTable["ham", names(counts.ham)] = counts.ham + .5  
    
    
    # Find the total number of spam and ham
    numSpam = sum(spam)
    numHam = length(spam) - numSpam
    
    # Prob(word|spam) and Prob(word | ham)
    wordTable["spam", ] = wordTable["spam", ]/(numSpam + .5)
    wordTable["ham", ] = wordTable["ham", ]/(numHam + .5)
    
    # log odds
    wordTable["presentLogOdds", ] = 
      log(wordTable["spam",]) - log(wordTable["ham", ])
    wordTable["absentLogOdds", ] = 
      log((1 - wordTable["spam", ])) - log((1 -wordTable["ham", ]))
    
    invisible(wordTable)
  }

# Obtain the probabilities and log odds for the training data.
trainTable = computeFreqs(trainMsgWords, trainIsSpam)

computeMsgLLR = function(words, freqTable) 
{
  # Discards words not in training data.
  words = words[!is.na(match(words, colnames(freqTable)))]
  
  # Find which words are present
  present = colnames(freqTable) %in% words
  
  sum(freqTable["presentLogOdds", present]) +
    sum(freqTable["absentLogOdds", !present])
}


##############################################################################################################



# Message text for testing removal of URLs in Exercise 1
testmsgBody=c("This is a url http://stat.illinois.edu.",
              "Sometimes URLs just start with a world wide web address like www.google.com.",
              "They can also be secure, like https://whatever.com.",
              "There are also file transfer protocol addresses like ftp://someagency.gov or ftps://filelocation.org.",
              "But we wouldn't want to remove www, http, https, ftp, or ftps on their own.")



```

```{r}

#A function to drop URLs from the message body
removeurls = function(messagebody){
    output = gsub("\\s?(((f|ht)(tp)s?(://))|www\\.)(.*)[.][a-z]+", "", messagebody)
              return(output)
}


#Testing the function on testmsgBody
removeurls(testmsgBody)

```

2. Write a function to extract the words from the subject value in a message header. The function should take a message header and stop words as its inputs. To implement this function, you can use the findMsgWordsfunction from the text inside your new function after doing some initial processing to get the subject content. Be sure to handle the case where no subject key is present in the message. Apply your function to the 15 sample email headers from sampleEmail to show that it works. 

```{r}
#Get the findMsgWords from the Homework3 code
findMsgWords = 
  function(msg, stopWords) {
    if(is.null(msg))
      return(character())
    
    words = unique(unlist(strsplit(cleanText(msg), "[[:blank:]\t]+")))
    
    # drop empty and 1 letter words
    words = words[ nchar(words) > 1]
    words = words[ !( words %in% stopWords) ]
    invisible(words)
  }

#Obtain the headerlist
headerList = lapply(sampleSplit, function(msg) msg$header)

#########################Function to extract subject words###################################################
# Get and see the subject locations in the headerlist and replace missing 'Subject' with NA.
# Use findMsgWords function to extract the words from the subject content

custom_func = function(headerList){
Subjectlines = sapply(headerList, function(header) {
  
  Subjectloc = grep("Subject", header)
  if (length(Subjectloc) == 0) return(NA)
  header[Subjectloc]
  
})

Subject_words = gsub("Subject:", "", Subjectlines)
Subject_words = findMsgWords(Subject_words, stopWords)
Subject_words
} 

#Test the function
custom_func(headerList)

```
3. 
```{r}
#Using the original function to find the log odds
computeFreqs =
  function(wordsList, spam, bow = unique(unlist(wordsList)))
  {
    # create a matrix for spam, ham, and log odds
    wordTable = matrix(0.5, nrow = 4, ncol = length(bow), 
                       dimnames = list(c("spam", "ham", 
                                         "presentLogOdds", 
                                         "absentLogOdds"),  bow))
    
    # For each spam message, add 1/2 to counts for words in message
    counts.spam = table(unlist(lapply(wordsList[spam], unique)))
    wordTable["spam", names(counts.spam)] = counts.spam + .5
    
    # Similarly for ham messages
    counts.ham = table(unlist(lapply(wordsList[!spam], unique)))  
    wordTable["ham", names(counts.ham)] = counts.ham + .5  
    
    
    # Find the total number of spam and ham
    numSpam = sum(spam)
    numHam = length(spam) - numSpam
    
    # Prob(word|spam) and Prob(word | ham)
    wordTable["spam", ] = wordTable["spam", ]/(numSpam + .5)
    wordTable["ham", ] = wordTable["ham", ]/(numHam + .5)
    
    # log odds
    wordTable["presentLogOdds", ] = 
      log(wordTable["spam",]) - log(wordTable["ham", ])
    wordTable["absentLogOdds", ] = 
      log((1 - wordTable["spam", ])) - log((1 -wordTable["ham", ]))
    
    invisible(wordTable)
  }

# Obtain the probabilities and log odds for the training data.
trainTable = computeFreqs(trainMsgWords, trainIsSpam)

computeMsgLLR = function(words, freqTable) 
{
  # Discards words not in training data.
  words = words[!is.na(match(words, colnames(freqTable)))]
  
  # Find which words are present
  present = colnames(freqTable) %in% words
  
  sum(freqTable["presentLogOdds", present]) +
    sum(freqTable["absentLogOdds", !present])
}

#Creating the train table
trainTable = computeFreqs(trainMsgWords, trainIsSpam)

#Obtaining the final log liklihood ratio of the test words
testLLR = sapply(testMsgWords, computeMsgLLR, trainTable)

#Obtaining the time taken to run the original function
system.time(sapply(testMsgWords, computeMsgLLR, trainTable))
```

```{r}

#Computing the log odds using the alternate formula

computeFreqs_alt =
  function(wordsList, spam, bow = unique(unlist(wordsList)))
  {
    # create a matrix for spam, ham, and log odds
    wordTable = matrix(0.5, nrow = 4, ncol = length(bow), 
                       dimnames = list(c("spam", "ham", 
                                         "presentLogOdds", 
                                         "absentLogOdds"),  bow))
    
    # For each spam message, add 1/2 to counts for words in message
    counts.spam = table(unlist(lapply(wordsList[spam], unique)))
    wordTable["spam", names(counts.spam)] = counts.spam + .5
    
    # Similarly for ham messages
    counts.ham = table(unlist(lapply(wordsList[!spam], unique)))  
    wordTable["ham", names(counts.ham)] = counts.ham + .5  
    
    
    # Find the total number of spam and ham
    numSpam = sum(spam)
    numHam = length(spam) - numSpam
    
    # Prob(word|spam) and Prob(word | ham)
    wordTable["spam", ] = wordTable["spam", ]/(numSpam + .5)
    wordTable["ham", ] = wordTable["ham", ]/(numHam + .5)
    
    # Alternate formula
    wordTable["presentLogOdds", ] = 
    wordTable["spam",]/wordTable["ham", ]
    wordTable["absentLogOdds", ] = 
    (1 - wordTable["spam", ])/(1 -wordTable["ham", ])
    
    invisible(wordTable)
  }


computeMsgLLR_alt = function(words, freqTable) 
{
  # Discards words not in training data.
  words = words[!is.na(match(words, colnames(freqTable)))]
  
  # Find which words are present
  present = colnames(freqTable) %in% words
  
  log(prod(freqTable["presentLogOdds", present])) +
    log(prod(freqTable["absentLogOdds", !present]))
}

#Computing the log odds ratio of the test cases using the above functions
trainTable_alt = computeFreqs_alt(trainMsgWords, trainIsSpam)
testLLR_alt = sapply(testMsgWords, computeMsgLLR_alt, trainTable_alt)

#Calculating the time taken to run this function

system.time(sapply(testMsgWords, computeMsgLLR_alt, trainTable_alt))


```

The time taken for the user defined function is higher.

```{r}
###############################################Comparing the results from both the functions###########################################################
testLLR = round(testLLR,3)
testLLR_alt = round(testLLR_alt,3)

#Let us get the indeces which are a mismatch
compare_output = is.na(match(testLLR, testLLR_alt))

Index_false = which(compare_output == "TRUE", arr.ind = TRUE)
Index_false
```
The above are the indeces which donot match because of the Inf values. Although mathematically on paper the same, the alternate function using the alternate formula gets Inf in a couple of instances.

We can see that both the results are not the same. There are a total of 29 mismatches and those are mentioned above. 

4.

```{r}

#The function given in the homework code

dropAttach = function(body, boundary){
  
  bString = paste("--", boundary, sep = "")
  bStringLocs = which(bString == body)
  
  # if there are fewer than 2 beginning boundary strings, 
  # there is on attachment to drop
  if (length(bStringLocs) <= 1) return(body)
  
  # do ending string processing
  eString = paste("--", boundary, "--", sep = "")
  eStringLoc = which(eString == body)
  
  # if no ending boundary string, grab contents between the first 
  # two beginning boundary strings as the message body
  if (length(eStringLoc) == 0) 
    return(body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1)])
  
  # typical case of well-formed email with attachments
  # grab contents between first two beginning boundary strings and 
  # add lines after ending boundary string
  n = length(body)
  if (eStringLoc < n) 
    return( body[ c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), 
                     ( (eStringLoc + 1) : n )) ] )

  return( body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1) ])
}

```

```{r}
#The alternate function for dropping the attachments when a third parameter is mentioned
#Code with the third parameter
dropAttach_alt = function(body, boundary, drop){
  
  bString = paste("--", boundary, sep = "")
  bStringLocs = which(bString == body)
  
  # do ending string processing
  eString = paste("--", boundary, "--", sep = "")
  eStringLoc = which(eString == body)
  
  # if there are fewer than 2 beginning boundary strings, 
  # there is no attachment to drop

  if (length(bStringLocs) <= 1 ){return(body)}
  
  
  if(drop == FALSE){
    
  body1 = body[ (bStringLocs[2] + 1)  : (eStringLoc - 1) ]
    
  
  Contentlines = sapply(body1, function(body) {
  Contentloc = grep("Content-Type", body)
  if (length(Contentloc) == 0) return(NA)
  body[Contentloc]
  })
  names(Contentlines) = NULL
  a = grep("(plain|html)", Contentlines )
  if (length(a) > 0)
  {
    n = length(body)
    # if no ending boundary string, grab contents between the first 
    # two beginning boundary strings as the message body
    if (length(eStringLoc) == 0) 
      return(body[c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), (bStringLocs[2] + 1) : n )])
    #typical case of well-formed email with attachments grab contents between first two beginning boundary strings 
    #and add lines after ending boundary string

    if (eStringLoc < n) 
    return( body[ c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), (bStringLocs[2] + 1)  : (eStringLoc - 1),
                     ( (eStringLoc + 1) : n )) ] )
  
  }
  }
  
  
  
  # if no ending boundary string, grab contents between the first 
  # two beginning boundary strings as the message body
  if (length(eStringLoc) == 0) 
    return(body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1)])
  
  # typical case of well-formed email with attachments
  # grab contents between first two beginning boundary strings and 
  # add lines after ending boundary string
  n = length(body)
  if (eStringLoc < n) 
    return( body[ c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), 
                     ( (eStringLoc + 1) : n )) ] )
  
  # fall through case
  # note that the result is the same as the 
  # length(eStringLoc) == 0 case, so code could be simplified by 
  # dropping that case and modifying the eStringLoc < n check to 
  # be 0 < eStringLoc < n
  return( body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1) ])
  
  }

#Calculating the different parameters that need to be passed into the function
bodyList = lapply(sampleSplit, function(msg) msg$body)
headerList = lapply(sampleSplit, function(msg) msg$header)
getBoundary = function(header) {
  boundaryIdx = grep("boundary=", header)
  boundary = gsub('"', "", header[boundaryIdx])
  gsub(".*boundary= *([^;]*);?.*", "\\1", boundary)
}
boundaryList = lapply(headerList, getBoundary)

#Result of the original function from homework code

result = mapply(dropAttach, bodyList, boundaryList)
lengths_1 = sapply(result, length) 
names(lengths_1) = NULL
print(lengths_1)

#Result when the third parameter is TRUE i.e 'drop attachments' 
result_alt_true = mapply(dropAttach_alt, bodyList, boundaryList, TRUE)
lengths_2 = sapply(result_alt_true, length)
names(lengths_2) = NULL
print(lengths_2)

#Result when the third parameter is FALSE i.e 'donot drop attachments'
result_alt_false = mapply(dropAttach_alt, bodyList, boundaryList, FALSE)
lengths_3 = sapply(result_alt_false, length)
names(lengths_3) = NULL
print(lengths_3)



```

As we can see the lengths are the same when the drop= TRUE but the lengths do differ when the drop= FALSE where the content type is plain text or HTML.
